---
title: "CAMBIUM_Model_Interface"
author: "Nemutshili Ndivho"
date: "2024-09-30"
output: html_document
---
## Introduction

This Shiny app provides an interface for running the **CAMBIUM Model**, a process-based model for simulating daily xylem development in Eucalyptus trees. Users can upload their custom input data, modify simulation parameters, and visualize results. The app is designed with multiple tabs to manage inputs, run the model, and display results, making it easier to use for researchers and analysts.

## Setup: Load Required Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load required libraries for the application
library(shiny)
library(shinythemes)
library(CAMBIUM)
library(DT)
library(shinyjs)
```

## Custom CSS for Styling

This section defines custom CSS that is applied throughout the Shiny app for consistent styling. It adjusts the background colors, button styles, headings, sidebar, and more for a polished user interface.

```{r Custom}
custom_css <- "
  body {
    background-color: #f0f8ff;
    font-family: Arial, sans-serif;
  }
  .nav-tabs {
    background-color: #009688;
  }
  .nav-tabs > li > a {
    color: white;
    padding: 15px;
  }
  .nav-tabs > li > a:hover {
    background-color: #00796b;
  }
  .btn-primary {
    background-color: #ff5722;
    border-color: #ff5722;
  }
  .btn-primary:hover {
    background-color: #e64a19;
    border-color: #e64a19;
  }
  h2, h4 {
    color: #004d40;
  }
  .dataTables_wrapper {
    margin: 20px;
  }
  .shiny-input-container {
    margin-bottom: 15px;
  }
  .plot-container {
    margin-top: 20px;
  }
  .sidebar {
    background-color: #009688;
    color: #fff;
  }
  .sidebar h3, .sidebar h4, .sidebar label {
    color: #fff;
  }
  .main-panel {
    background-color: #ffffff;
    padding: 20px;
    border-radius: 8px;
    box-shadow: 0px 0px 15px rgba(0, 0, 0, 0.1);
  }
  .footer {
    text-align: center;
    padding: 20px;
    font-size: 12px;
    color: #666;
    background-color: #f0f0f0;
  }
  .loading-spinner {
    display: none;
    text-align: center;
  }
  .loading-spinner img {
    width: 50px;
  }
"
```

## User Interface (UI)

This section defines the User Interface (UI) for the Shiny app. It creates multiple tabs including the Home, Run Model, Parameters & Instructions, and About tabs. Each tab contains specific controls and displays, making the interface user-friendly and functional.

```{r UI}
# Define UI for Application
ui <- fluidPage(
  theme = shinytheme("cerulean"),

  # Include the custom CSS
  tags$head(
    tags$style(HTML(custom_css))
  ),

  # Define the navigation bar with different tabs
  navbarPage(
    "CAMBIUM Model Interface",

    # Home Tab
    tabPanel("Home",
             fluidRow(
               column(12,
                      div(style = "text-align: center; padding: 50px;",
                          h2("Welcome to the CAMBIUM Model Shiny App"),
                          h4("Process-Based Model of Daily Xylem Development in Eucalyptus"),
                          p("This application allows you to simulate daily xylem development in Eucalyptus trees using the CAMBIUM model. Customize inputs, run simulations, and visualize results with ease."),
                          br(),
                          h4("Key Features:"),
                          tags$ul(
                            tags$li("Upload custom forcing and parameter data"),
                            tags$li("Adjust simulation settings"),
                            tags$li("View and edit model parameters")
                          ),
                          br(),
                          h4("Use the navigation bar above to get started!")
                      )
               )
             )
    ),

    # Run Model Tab
    tabPanel("Run Model",
             sidebarLayout(
               sidebarPanel(
                 class = "sidebar",
                 h3("Input Data"),
                 fileInput("file_forcing", "Upload Forcing Data (CSV)", accept = c(".csv")),
                 fileInput("file_param", "Upload Parameters Data (CSV)", accept = c(".csv")),
                 h3("Settings"),
                 checkboxInput("debug_print", "Enable Debug Printing", value = TRUE),
                 actionButton("run_model", "Run CAMBIUM Model", class = "btn btn-primary"),
                 downloadButton("download_results", "Download Results as CSV", class = "btn btn-primary")
               ),

               mainPanel(
                 class = "main-panel",
                 h4("Model Status"),
                 textOutput("run_status"),
                 div(class = "loading-spinner", id = "loading_spinner",
                     img(src = "https://www.wpfaster.org/wp-content/uploads/2013/06/loading-gif.gif"),
                     p("Running... Please wait.")
                 ),
                 br(),
                 h4("Model Output"),
                 verbatimTextOutput("model_output")
               )
             )
    ),

    # Parameters & Instructions Tab
    tabPanel("Parameters & Instructions",
             sidebarLayout(
               sidebarPanel(
                 class = "sidebar",
                 h4("Model Parameters")
               ),

               mainPanel(
                 class = "main-panel",
                 h4("Parameters Table"),
                 DTOutput("params_table"),
                 br(),
                 h4("Instructions Table"),
                 DTOutput("instructions_table")
               )
             )
    ),

    # About Tab
    tabPanel("About",
             fluidRow(
               column(12,
                      div(style = "text-align: center; padding: 50px;",
                          h3("About CAMBIUM"),
                          p("The CAMBIUM model simulates daily xylem development in Eucalyptus trees. This app provides an interface for running the model with user-provided data."),
                          br(),
                          h4("References"),
                          p("Drew, D. M., Downes, G. M., & Battaglia, M. (2010). CAMBIUM, a process-based model of daily xylem development in Eucalyptus. Journal of Theoretical Biology, 264(2), 395–406. https://doi.org/10.1016/j.jtbi.2010.02.013"),
                          br(),
                          h4("Created by Nemutshili Ndivho Confidence")
                      )
               )
             )
    )
  ),

  # Footer
  tags$footer(class = "footer",
              "© 2024 Nemutshili Ndivho Confidence. All rights reserved."
  )
)
```

## Server Logic

The server logic processes the input files, runs the CAMBIUM model, and displays the results. It includes data validation, error handling, and displays the model output on the UI.

```{r Server}
server <- function(input, output, session) {

  # ---- Reactive Values ----
  model_results <- reactiveVal(NULL)

  # ---- Render Data Tables ----
  output$params_table <- renderDT({
    data(params)  # Ensure 'params' is a valid data frame
    datatable(params, options = list(pageLength = 10, autoWidth = TRUE))
  })

  output$instructions_table <- renderDT({
    data(runsettings)  # Ensure 'runsettings' is a valid data frame
    datatable(runsettings, options = list(pageLength = 10, autoWidth = TRUE))
  })

  # ---- Event Handlers ----
  observeEvent(input$run_model, {
    req(input$file_forcing, input$file_param)

    # Show loading spinner
    shinyjs::show("loading_spinner")

    start_time <- Sys.time()

    # Read the uploaded files
    filename_forcing <- input$file_forcing$datapath
    filename_param <- input$file_param$datapath

    # Run the CAMBIUM model
    output$run_status <- renderText("Running CAMBIUM Model...")

    tryCatch({
      if (input$debug_print) {  # Correctly using input$debug_print
        # Print debug information before running the model
        cat("Debug: Forcing file path:", filename_forcing, "\n")
        cat("Debug: Parameter file path:", filename_param, "\n")
        cat("Debug: Run settings:\n")
        str(runsettings)  # Use str() to display the structure of runsettings
      }

      # Capture the results from the CAMBIUM model run
      results <- run_CAMBIUM(
        filename_forcing = filename_forcing,
        filename_param = filename_param,
        run_settings = runsettings
      )

      model_results(results)  # Store results in a reactive value
      output$model_output <- renderText("Model run completed successfully.")

      # Hide loading spinner and display run duration
      shinyjs::hide("loading_spinner")
      run_duration <- Sys.time() - start_time
      output$run_status <- renderText(paste("CAMBIUM Run duration was:", round(run_duration, 2), "seconds."))

    }, error = function(e) {
      output$model_output <- renderText(paste("Error in running CAMBIUM model:", e$message))

      # Hide loading spinner in case of error
      shinyjs::hide("loading_spinner")
    })
  })

  # ---- Download Handler for CSV ----
  output$download_results <- downloadHandler(
    filename = function() {
      paste("CAMBIUM_results_", Sys.Date(), ".csv", sep = "")
    },
    content = function(file) {
      # Get the list of objects in the environment
      object_names <- ls(envir = .GlobalEnv)

      # Create an empty list to store the objects as columns
      columns <- list()

      # Find the maximum length of all objects that have more than one value
      max_length <- max(sapply(object_names, function(x) {
        obj_len <- length(get(x, envir = .GlobalEnv))
        if (obj_len > 1) return(obj_len)
        return(0)  # Ignore objects with only one value
      }))

      # Loop through each object and store it as a column in the list
      for (object_name in object_names) {
        obj <- get(object_name, envir = .GlobalEnv)  # Retrieve the object from the global environment

        # Ensure the object is of a compatible type (numeric, character, logical)
        if ((is.numeric(obj) || is.character(obj) || is.logical(obj)) && length(obj) > 1) {
          # Pad the object with NAs to make its length equal to max_length
          padded_obj <- c(obj, rep(NA, max_length - length(obj)))
          columns[[object_name]] <- padded_obj
        } else {
          cat("Skipping", object_name, "- not a compatible type (numeric/character/logical) or has only one value\n")
        }
      }

      # Convert the list to a data frame named 'Output'
      if (length(columns) > 0) {
        Output <- as.data.frame(columns)

        # Write the output to the CSV file
        write.csv(Output, file = file, row.names = FALSE)
      } else {
        cat("No compatible objects with more than one value found\n")
      }
    }
  )
}
```

## Run the Application

The Shiny app is launched using the shinyApp function, combining both the UI and server components.

```{r Run App}
shinyApp(ui = ui, server = server)
```
